---
title: "p8105_hw2_yw3996"
author: "Yiying Wu"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load R packages
```{r,include=TRUE,message = FALSE}
library(tidyverse)
```


## Problem 1: FiveThirtyEight data
Clean the data in `pols-month.csv`: 

* Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day` 
* Replace month number with month name
* Create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop` 
* Remove the `day` variable.
```{r}
pols_month = read_csv("./fivethirtyeight_datasets/pols-month.csv")|>
  janitor::clean_names()|>
  separate(mon,into=c("year","month","day"),)|>
  mutate(month = month.name[as.numeric(month)],
         president=case_match(
           prez_dem,
           1~"democratic",
           0~"republican"
           )
        )|>
  select(-prez_dem,-prez_gop,-day)

pols_month  
```

Clean the data in snp.csv using a similar process to the above, arrange according to year and month, and organize so that year and month are the leading columns.
```{r}
snp = read_csv("./fivethirtyeight_datasets/snp.csv")|>
  janitor::clean_names()
```



