---
title: "p8105_hw2_yw3996"
author: "Yiying Wu"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load R packages
```{r,include=TRUE,message = FALSE}
library(tidyverse)
library(readxl)
```


## Problem 1: FiveThirtyEight data
### `pols-month.csv`
Clean the data in `pols-month.csv`: 

* Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day` 
* Replace month number with month name
* Create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop` 
* Remove the `day` variable.
```{r}
pols_month = read_csv("./fivethirtyeight_datasets/pols-month.csv")|>
  janitor::clean_names()|>
  separate(mon,into=c("year","month","day"),)|>
  mutate(month = month.name[as.numeric(month)],
         president=case_match(
           prez_dem,
           1~"democratic",
           0~"republican"
           )
        )|>
  select(-prez_dem,-prez_gop,-day)

pols_month
```

### `snp.csv`
Clean the data in `snp.csv` using a similar process to the above. 

* arrange according to year and month
* organize so that year and month are the leading columns
```{r}
snp = read_csv("./fivethirtyeight_datasets/snp.csv")|>
  janitor::clean_names()|>
  mutate(
    date=ifelse(as.Date(date,"%m/ %d/ %y")> Sys.Date(),
                format(as.Date(date,"%m/ %d/ %y"),'19%y-%m-%d'),
                format(as.Date(date,"%m/ %d/ %y"),'%Y-%m-%d'))
        )|>
  separate(date,into=c("year","month","day"),)|>
  mutate(month = month.name[as.numeric(month)])|>
  select(year,month, everything(),-day)

snp
```

### `unemployment`
tidy the `unemployment` data so that it can be merged with the previous datasets: 

* switching from “wide” to “long” format
* ensuring that key variables have the same name
* ensuring that key variables take the same values
```{r}
unemployment = read_csv("./fivethirtyeight_datasets/unemployment.csv")|>
  pivot_longer(
    Jan:Dec,
    names_to = "month", 
    values_to = "unemployment")|>
  janitor::clean_names()|>
  mutate(month=month.name[match(month,month.abb)])|>
  mutate(year=as.character(year))

unemployment
```

### Merge datasets
Join the datasets by merging `snp` into `pols_month`, and merging `unemployment` into the result.
```{r}
df=pols_month|>
  left_join(snp,by=c("year", "month"))|>
  left_join(unemployment, by=c("year", "month"))

df
```

### Description of datasets
* `pols_month`: pols_month is a `r nrow(pols_month)` rows × `r ncol(pols_month)` columns dataset with variable `year`, `month`, `gov_gop`, `sen_gop`, `rep_gop`, `gov_dem`, `sen_dem`, `rep_dem`, `president`. The `year` variable in this dataset ranges from `r min(pols_month$year)` to `r max(pols_month$year)`.
* `snp`: snp is a `r nrow(snp)` rows × `r ncol(snp)` columns dataset with variable `year`, `month`, `close`. The `year` variable in this dataset ranges from `r min(snp$year)` to `r max(snp$year)`.
* `unemployment`: unemployment is a `r nrow(unemployment)` rows × `r ncol(unemployment)` columns dataset with variable `year`, `month`, `unemployment`. The `year` variable in this dataset ranges from `r min(unemployment$year)` to `r max(unemployment$year)`.
* `df`: the joined dataset df is a `r nrow(df)` rows × `r ncol(df)` columns dataset with variable `year`, `month`, `gov_gop`, `sen_gop`, `rep_gop`, `gov_dem`, `sen_dem`, `rep_dem`, `president`, `close`, `unemployment`. The `year` variable in this dataset ranges from `r min(df$year)` to `r max(df$year)`.


## Problem 2: Mr. Trash Wheel dataset
### Mr. Trash Wheel sheet
Read and clean the Mr. Trash Wheel sheet:

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in `read_excel`
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* update the data to include a new homes_powered variable based on Homes Powered - Each ton of trash equates to on average 500 kilowatts of electricity.  An average household will use 30 kilowatts per day..
```{r}
mr_trash_wheel = read_excel(
  "202207 Trash Wheel Collection Data.xlsx", 
  sheet="Mr. Trash Wheel", range = cell_cols("A:N"))|>
  janitor::clean_names()|>
  filter(row_number() <= n()-1)|>
  mutate(homes_powered=weight_tons*500/30)

mr_trash_wheel
```

### Professor Trash Wheel sheet
Use a similar process to import, clean, and organize the data for `professor_trash_wheel`
```{r}
professor_trash_wheel = read_excel(
  "202207 Trash Wheel Collection Data.xlsx", 
  sheet="Professor Trash Wheel", range = cell_cols("A:M"))|>
  janitor::clean_names()|>
  filter(row_number() <= n()-1)|>
  mutate(homes_powered=weight_tons*500/30)

professor_trash_wheel
```

### Gwynnda Trash Wheel sheet
Use a similar process to import, clean, and organize the data for `gwynnda`
```{r}
gwynnda = read_excel(
  "202207 Trash Wheel Collection Data.xlsx", 
  sheet="Gwynnda Trash Wheel", range = cell_cols("A:K"))|>
  janitor::clean_names()|>
  filter(row_number() <= n()-2)|>
  mutate(homes_powered=weight_tons*500/30)

gwynnda
```

### Combine datasets
combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset `mr_trash_wheel_tidy`. To keep track of which Trash Wheel is which, add an additional variable `trash_wheel` to all datasets before combining.
```{r}
mr_trash_wheel=mr_trash_wheel|>
  mutate(trash_wheel="Mr. Trash Wheel")

professor_trash_wheel=professor_trash_wheel|>
  mutate(trash_wheel="Professor Trash Wheel",
         year=as.character(year))

gwynnda=gwynnda|>
  mutate(trash_wheel="Gwynnda Trash Wheel",
         year=as.character(year))

mr_trash_wheel_tidy = 
  bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda) |>
  janitor::clean_names() |>
  select(trash_wheel, everything()) 

mr_trash_wheel_tidy
```

### Description of datasets

* The resulting dataset `mr_trash_wheel_tidy` includes `r nrow(mr_trash_wheel_tidy)` observations, with `r ncol(mr_trash_wheel_tidy)` variables: `trash_wheel`, `dumpster`, `month`, `year`, `date`, `weight_tons`, `volume_cubic_yards`, `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `grocery_bags`, `chip_bags`, `sports_balls`, `homes_powered`, `plastic_bags`. 
* Key variables
  * `trash_wheel`: Indicate which sheet each record belongs to. "Mr. Trash Wheel" means the record is in Mr. Trash Wheel sheet. "Professor Trash Wheel" means the record is in Professor Trash Wheel sheet. "Gwynnda Trash Wheel" means the record is in Gwynnda Trash Wheel sheet.
  * `homes_powered`: Homes Powered - Each ton of trash equates to on average 500 kilowatts of electricity.  An average household will use 30 kilowatts per day. This can be calculates as weight_tons * 500 / 30
* The total weight of trash collected by Professor Trash Wheel is `r sum(professor_trash_wheel$weight_tons)` tons.
* The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(subset(gwynnda, month == "July" & year == "2021" )$cigarette_butts)`.

## Problem 3: MCI data
### Baseline Demographics
Import, clean, and tidy the dataset of baseline demographics

* Ensure sex and APOE4 carrier status are appropriate encoded (i.e. not numeric)
* Remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline)
```{r}
mci_baseline = read_csv("./data_mci/MCI_baseline.csv", skip=1)|>
  janitor::clean_names()|>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        0 ~ "female"),
    sex = as.factor(sex),
    apoe4 = 
      case_match(
        apoe4,
        1 ~ "apoe4 carrier",
        0 ~ "apoe4 non-carrier"),
    apoe4 = as.factor(apoe4),
    age_at_onset = ifelse(age_at_onset==".",NA,age_at_onset)
    ) |>
  drop_na(age_at_onset)

mci_baseline
```

Discuss important steps in the import process and relevant features of the dataset.

```{r}
mci_amyloid = read_csv("./data_mci/mci_amyloid.csv", skip=1)|>
  janitor::clean_names()


```

